---
title: "Homework 2"
author: "Julia Thompson"
date: "10/1/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
```

## Problem 1

Read and cleaned Mr. Trash Wheel data.

```{r}
trash_data = read_excel(
  path = "./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
  sheet = "Mr. Trash Wheel",
  range = "A2:N408") %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(sports_balls = as.integer(sports_balls))
```

Read and cleaned precipitation data for 2017 and 2018. Rows without precipitation data were omitted and a variable was added for _year_.

```{r}
precip_2017 = read_excel(
  path = "./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
  sheet = "2017 Precipitation",
  skip = 1) %>% 
  janitor::clean_names() %>% 
  drop_na(month) %>% 
  mutate(year = 2017)

precip_2018 = read_excel(
  path = "./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
  sheet = "2018 Precipitation",
  skip = 1) %>% 
  janitor::clean_names() %>% 
  drop_na(month) %>% 
  mutate(year = 2018)
```

Combined precipitation datasets and converted month to a character variable.

```{r}

# CHANGE THIS AND USE ONE OF THE JOIN FUNCITONS!

precip_data = 
  bind_rows(precip_2017, precip_2018) %>% 
  pivot_wider(
  names_from = year,
  values_from = total) %>% 
  rename("total_precip_2017" = "2017", "total_precip_2018" = "2018") %>% 
  mutate(
    month = month.name[]
  )
```

The first dataset created was _trash_data_, which is made up of `r nrow(trash_data)` observations and `r ncol(trash_data)` variables. It contains a variable _dumpster_, which ranges from `r min(pull(trash_data,dumpster))` to `r max(pull(trash_data,dumpster))` and denotes the dumpster ID. For each dumpster, we have variables _month_, _year_, and _date_ to inform us about when the data was collected. Additionally, there are several variables that correspond to specifics about what each dumpster collected (such as _plastic_bottles_ and _chip_bags_). The median number of sports balls collected in a dumpster in 2017 was `r median(pull(trash_data, sports_balls)[pull(trash_data, year)==2017])`.

The second dataset created was _precip_data_, made up of `r nrow(precip_data)` observations and `r ncol(precip_data)` variables. The variables are _month_, _total_precip_2017_, and _total_precip_2018_, which give the month of the year (January through December), and inches of rainfall per month for 2017 and 2018 respectively. The total precipitation in 2018 was `r sum(pull(precip_data, total_precip_2018))` inches, which is a large increase compared to the total precipitation in 2017 of `r sum(pull(precip_data, total_precip_2017))` inches.

## Problem 2

First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

```{r}
pols_data = read_csv(
  file = "./data/pols-month.csv") %>% 
  separate(col = mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) %>% 
  mutate(
    prez_gop = ceiling(prez_gop/2),
    month = month.name[month],
    president = prez_gop - prez_dem,
    president = recode(president, `-1` = "dem", `1` = "gop")
  ) %>% 
  select(year, month, president, everything(), -day)
```

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r}
snp_data = read_csv(
  file = "./data/snp.csv") %>% 
  separate(col = date, into = c("month", "day", "year"), sep = "/", convert = TRUE) %>% 
  mutate(
    month = month.name[month]
  ) %>% 
  select(year, month, everything(), -day)
```

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

