---
title: "Homework 2"
author: "Julia Thompson"
date: "10/1/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)
```

## Problem 1

Read and cleaned Mr. Trash Wheel data.

```{r}
trash_data = read_excel(
  path = "./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
  sheet = "Mr. Trash Wheel",
  range = "A2:N408") %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(sports_balls = as.integer(sports_balls))
```

Read and cleaned precipitation data for 2017 and 2018. Rows without precipitation data were omitted and a variable was added for _year_.

```{r}
precip_2017 = read_excel(
  path = "./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
  sheet = "2017 Precipitation",
  skip = 1) %>% 
  janitor::clean_names() %>% 
  drop_na(month) %>% 
  mutate(year = 2017)

precip_2018 = read_excel(
  path = "./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx",
  sheet = "2018 Precipitation",
  skip = 1) %>% 
  janitor::clean_names() %>% 
  drop_na(month) %>% 
  mutate(year = 2018)
```

Combined precipitation datasets and converted month to a character variable.

```{r}
precip_data = 
  bind_rows(precip_2017, precip_2018) %>% 
  mutate(
    month = month.name[month]
  )
```

The first dataset created was _trash_data_, which is made up of `r nrow(trash_data)` observations and `r ncol(trash_data)` variables. It contains a variable _dumpster_, which ranges from `r min(pull(trash_data,dumpster))` to `r max(pull(trash_data,dumpster))` and denotes the dumpster ID. For each dumpster, we have variables _month_, _year_, and _date_ to inform us about when the data was collected. Additionally, there are several variables that correspond to specifics about what each dumpster collected (such as _plastic_bottles_ and _chip_bags_). The median number of sports balls collected in a dumpster in 2017 was `r median(pull(trash_data, sports_balls)[pull(trash_data, year)==2017])`.

The second dataset created was _precip_data_, made up of `r nrow(precip_data)` observations and `r ncol(precip_data)` variables. The variables are _month_, _total_, and _year_, which give the month of the year (January through December), total inches of rainfall, and year (2017 or 2018) respectively. The total precipitation in 2018 was `r sum(pull(precip_data, total)[pull(precip_data, year)==2018])` inches, which is a large increase compared to the total precipitation in 2017 of `r sum(pull(precip_data, total)[pull(precip_data, year)==2017])` inches.

## Problem 2

First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

Note: a few values in prez_gop had 2 instead of 1. Recoded those to be 1 below.

```{r}
pols_data = read_csv(
  file = "./data/pols-month.csv") %>% 
  separate(col = mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) %>% 
  mutate(
    prez_gop = ceiling(prez_gop/2),
    month = month.name[month],
    president = prez_gop - prez_dem,
    president = recode(president, `-1` = "dem", `1` = "gop")
  ) %>% 
  select(year, month, president, everything(), -day, -prez_gop, -prez_dem)
```

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r}
snp_data = read_csv(
  file = "./data/snp.csv") %>% 
  separate(col = date, into = c("month", "day", "year"), sep = "/", convert = TRUE) %>% 
  mutate(
    month = month.name[month]
  ) %>% 
  select(year, month, everything(), -day)
```

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

**This is horrible and there must be a better way. Either use rename(1:12 = jan:dec), or just use month.abb from the beginning and not get the full name.**

```{r}
unemp_data = read_csv(
  file = "./data/unemployment.csv") %>% 
  janitor::clean_names() %>% 
  rename(`1` = jan, `2` = feb, `3` = mar, `4` = apr, `5` = may, `6` = jun, `7` = jul, `8` = aug, `9` = sep,           `10` = oct, `11` = nov, `12` = dec) %>% 
  pivot_longer(
    cols = `1`:`12`,
    names_to = "month",
    values_to = "unemp_pct"
  ) %>% 
  mutate(
    month = as.integer(month),
    month = month.name[month]
  )

```

Join the datasets by merging snp into pols, and merging unemployment into the result.

```{r}
snp_pols = 
  left_join(pols_data, snp_data, by = c("year","month"))

snp_pols_unemp = 
  left_join(snp_pols, unemp_data, by = c("year", "month"))
```

## Problem 3

**Is there a better way to recode these?**

```{r}
baby_data = read_csv(
  file = "./data/Popular_Baby_Names.csv") %>% 
    janitor::clean_names() %>% 
  mutate(
    childs_first_name = str_to_sentence(childs_first_name),
    ethnicity = recode(ethnicity, "WHITE NON HISP" = "WHITE NON HISPANIC",
                                    "ASIAN AND PACI" = "ASIAN AND PACIFIC ISLANDER",
                                    "BLACK NON HISP" = "BLACK NON HISPANIC")
  ) %>% 
  distinct()
```

Produce a well-structured, reader-friendly table showing the rank in popularity of the name “Olivia” as a female baby name over time; this should have rows for ethnicities and columns for year. Produce a similar table showing the most popular name among male children over time.

```{r}
# we want rows to be ethnicities and columns to be years that hold rank

pop_olivia = filter(baby_data, childs_first_name == "Olivia" & gender == "FEMALE") %>% 
  pivot_wider(
    id_cols = c("year_of_birth", "ethnicity", "rank"),
    names_from = "year_of_birth",
    values_from = "rank"
  )

knitr::kable(pop_olivia)
```

Produce a similar table showing the most popular name among male children over time.

```{r}
pop_male = filter(baby_data, rank == "1" & gender == "MALE") %>% 
  pivot_wider(
    id_cols = c("year_of_birth", "ethnicity", "childs_first_name"),
    names_from = "year_of_birth",
    values_from = "childs_first_name"
  )

knitr::kable(pop_male)
```

Finally, for male, white non-hispanic children born in 2016, produce a scatter plot showing the number of children with a name (y axis) against the rank in popularity of that name

```{r}
#count vs rank for male white non hispanic children in 2016
white_male_2016 = filter(baby_data, gender == "MALE" & year_of_birth == "2016" & ethnicity == "WHITE NON HISPANIC") 

ggplot(white_male_2016, aes(x=rank, y=count)) + 
  geom_point() +
  labs(
    title = "Rank vs Count for Male White non-Hispanic Children born in 2016",
    x = "Rank",
    y = "Count"
  )
```

